### 2012-10-24 rnaseq_paper.tex
corrected some numbers, edited the intro

\section*{Introduction}
Until recently, studies of alternative splicing had been limited to a small number of genes and isoforms due to high-cost and low-throughput sequencing of expressed sequence tags (ESTs) and full-length cDNA libraries.  
RNA sequencing (RNA-Seq) using the next-generation sequencing (NGS) has been used successfully in many studies to gain unprecedented insight into a complexity of transcriptomes.
Novel isoforms have been identified based on reads mapped across known splice junctions or putative splice junctions from annotations.
It has been estimated that, in human, $92-94$\% of multiexon genes undergo alternative splicing and different isoforms are expressed in different tissues\cite{Wang:2008ea}.
This suggests that even in human a large number of splice variants have not been explored.

Sequences from RNA-Seq are very short (~75--250bp for Illumina reads); therefore, it is not feasible to assess gene
and isoform expression without mapping reads to a reference annotation or construct a full-length transcript from
\emph{de novo} assembly.
Several softwares employing different approaches have been developed to reconstruct transcripts from short reads.
Tophat\cite{Trapnell:2009dp} maps reads to a reference genome to identify exons and splice sites \emph{ab initio}.
It first searches for known splie sites in a genome and then tries to map reads across splice juntions to identify exon junctions.
Exon junctions can be used to build gene models by Cufflinks\cite{Trapnell:2010kd}.
On the other hand, Trinity\cite{Grabherr:2011jb}, has been developed to assemble short reads \emph{de novo} to construct transcripts.
Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are an extension of Abyss\cite{Simpson:2009iv} and Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} assembler that are designed to work with RNA-Seq data.
These programs are comparable at identifying alternative splicing with a slightly different sensitivity and specificity.
However, Oases with Oases-M has been shown to be superior than other \emph{de novo} assembler at discovering isoforms in human and mouse\cite{Schulz:2012je}.

In this study, we present a pipeline that uses Velvet and Oases assembler to discover alternative splicing in RNA-Seq data from chickens.
We use a new technique called a local assembly that enhances sensitivity of alternative splicing detection of Oases.
The results show that the pipeline can detect more isoforms than Oases-M and can detect isoforms not found by
Cufflinks.

@@

We developed an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach,
implemented in a software package called Gimme, merges transcripts and
gene models based on overlapping exons using an exon-graph approach
(see Materials and Methods; Figure XX).  We used Gimme to obtain
20,439 gene models containing 25,859 isoforms from our global
assembly; 21,963 gene models containing 25,859 from our local
assembly; and 20,922 gene models containing 27,732 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).

@@

Of 104,827 splice junctions from the gene models, 248 (0.23\%)
junctions have no spliced reads and only 484 (0.46\%) junctions
have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
% @CTB how do you get junctions with no spliced reads!?
More than 99\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?

@@

Of the 104,827 splice junctions in our gene models, 82,908 (80.31\%)
are supported by ESTs or mRNAs from Genbank.  This is especially
surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues.  Note that this
cross-validation suggests that the 19,981 novel splice junctions {\em not}
seen in publicly available ESTs and mRNAs are also likely to be real
splice junctions from spleen.
% @CTB supported by ESTs OR mRNAs, or ESTs AND mRNAs?  Surely the former?

@@

As shown above, Cufflinks and the assembly method detect a number of
distinct but equally valid splice junctions, which suggests that we
could obtain greater sensitivity to exon-exon junctions in our gene models by merging both sets of predictions.
We therefore used Gimme to merge the Cufflinks and assembly gene
models Table.~\ref{genes_transcripts}.  This resulted in a decreased
number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{cufflinks_asm}). The merged gene models recover 51\% and
66\% of splice junctions from ESTs and ENSEMBL respectively, which is
10-15\% greater than that from corresponding unmerged gene models.
% @CTB what about splice junctions etc??

@@

To validate our predicted isoforms, we extracted putative coding sequences from
our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan
successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
protein sequences with 50 or more amino acids.  We then searched for
homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
of our isoforms from 12,399 distinct genes match mouse proteins at a
bit score of XXX.
% @CTB
Most of these matches have a bit-score/length ratio greater than 1,
which indicates a good agreement between chicken and mouse proteins.
% @CTB why so few match, really?

@@

\caption{
\bf{Number of putative genes and isoforms}}
\begin{tabular}{cccccc}
\hline
Method& Gene & Isoform \\ 
\hline
Global & 20,439 & 26,035 \\
Local & 21,963 & 25,859 \\
Global + Local & 20,922 & 27,732 \\
Cufflinks & 25,318 & 34,959 \\
Global + Local + Cufflinks + mRNAs & 26,726 & 43,074\\
Ensembl & 17,934 & 23,392 \\
\hline
%table information
\end{tabular}
%\begin{flushleft}\footnotesize \textit{*Minumum set}
%\end{flushleft}
\label{genes_transcripts}

>>>>

\section*{Introduction}
Until recently, studies of alternative splicing had been limited to a small
number of genes and isoforms due to high-cost and low-throughput sequencing
of expressed sequence tags (ESTs) and full-length cDNA libraries.  
RNA sequencing (RNA-Seq) using the next-generation sequencing (NGS) has been
used successfully in many studies to gain unprecedented insight into a
complexity of transcriptomes.
It has been estimated that, in human, $92-94$\% of multiexon genes undergo alternative
splicing and different isoforms are expressed in different tissues\cite{Wang:2008ea}.
This suggests that even in human a large number of splice variants have not been explored.

%Since sequences from RNA-Seq (reads) are very short (~50--250bp for Illumina reads),
%it is nokt feasible to assess expression isoforms and their structures without mapping reads
%to a reference annotation or reconstructing a full-length transcript from short reads.

Despite the small size of sequencing reads, several studies have detected
novel splice junctions based on alignment of reads spanning across putative exon junctions.
To map reads across exon junctions, reads are split into two parts and each part is mapped
to the genome independently.
A splice junction is then identified based on alignments of each half of a read
that fall between two exons at exon-intron boundaries.
With this approach, Wang \emph{et al} have identified a large number of
splice junctions that are not annotated from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.
These novel splice sites include both canonical and non-canonical splice sites.
Approximately, $46\%-75\%$ of canonical splice sites are supported by ESTs.
Novel splice junctions have different levels of read coverage suggesting that both high-
and low-expressed isoforms are unannotated.

Using a similar approach, Pickrell \emph{et al} identified more than 150,000 novel canonical splice junctions
in lymphoblastoid cells.
The study also shows that the number of unannotated splice junctions varies among cells from
different human tissues, which suggests tissue-specific expression of isoforms\cite{Pickrell:2010gt}.
A majority of unannotated isoforms found in both studies are from alternative splicing
of annotated exons. Only a small fraction of isoforms contain exons from intergenic or
intronic region.

Several tools have been developed not only to detect novel splice junctions but also to reconstruct
full-length isoforms from short reads without using gene annotations.
These tools are especially useful for transcriptome analysis of organims with poor gene annotations.
Cufflinks\cite{Trapnell:2010kd} relies on splice junctions detected from Tophat\cite{Trapnell:2009dp},
a read aligner that can align reads across putative exon junctions, to reconstruct a full-length transcript.
Cufflinks identified 12,712 novel isoforms, of which 7,395 ($58\%$) contain novel splice junctions in
mouse myoblast cell lines.
Guttman \emph{et al} used Scripture, a tool employing similar mapping-based approach, to reconstruct
a full-length transcripts from mouse RNA-Seq data and discovered approximately 490 novel alternative
isoforms in lincRNA loci, which are expressed in any of the three different cell types\cite{Guttman:2010io}.

Although the mapping-based method has been useful in both detecting splice junctions and isoforms,
it relies heavily on a reference genome.
Hence, organisms lacking a genome or a high-quality one may not find this approach practical.
This limitation can be overcome by \emph{de novo} assembly of short reads.

A number of \emph{de novo} assemblers have been used to reconstruct transcripts from RNA-Seq data in
many studies.
Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct transcripts from yeast and mouse datasets.
It was also shown that Trinity detects a unique set of novel splice junctions not detected by Cufflinks
or Scripture.
This suggests that \emph{de novo} assembly approach is capabale of increasing sensitivity of detecting
alternative isoforms of a mapping-based method.
Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are an extension
of Abyss\cite{Simpson:2009iv} and Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assembler
that are tuned to work with RNA-Seq data.
These assemblers are comparable at reconstructing existing and novel alternative isoforms with a slightly
different sensitivity and specificity.
However, Oases with Oases-M has been shown to be superior than other \emph{de novo}
assemblers at discovering isoforms in human and mouse\cite{Schulz:2012je}.

In this study, we present a pipeline that uses Velvet and Oases assembler to
reconstruct alternative isoforms in RNA-Seq data from chickens.
We use a new technique called a local assembly that enhances sensitivity of
alternative isoform detection of Oases.
The results show that the pipeline can detect more isoforms than Oases-M and
can detect isoforms not found by Cufflinks.
We also showed that transcripts reconstructed from \emph{de novo} assembly and mapping-based approach can be
merged to build more complete predicted gene models.

@@

We developed an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach,
implemented in a software package called Gimme, merges transcripts and
gene models based on overlapping exons using an exon-graph approach
(see Materials and Methods; Figure~\ref{algorithm}).  We used Gimme to obtain
20,439 gene models containing 25,859 isoforms from our global
assembly; 21,963 gene models containing 25,859 from our local
assembly; and 20,922 gene models containing 27,732 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).

@@

Of 100,411 splice junctions from the gene models, 95 (0.09\%)
junctions have no spliced reads and only 448 (0.45\%) junctions
have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
% @CTB how do you get junctions with no spliced reads!?
More than 99\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?

@@

Of the 100,411 splice junctions in our gene models, 81,838 (81.50\%)
are supported by ESTs or mRNAs from Genbank.  This is especially
surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues.  Note that this
cross-validation suggests that the 18,573 novel splice junctions {\em not}
seen in publicly available ESTs and mRNAs are also likely to be real
splice junctions from spleen.
% @CTB supported by ESTs OR mRNAs, or ESTs AND mRNAs?  Surely the former?

@@

As shown above, Cufflinks and the assembly method detect a number of
distinct but equally valid splice junctions, which suggests that we
could obtain greater sensitivity to exon-exon junctions in our gene
models by merging both sets of predictions.
We therefore used Gimme to merge the Cufflinks and assembly gene
models Table.~\ref{genes_transcripts}.  This resulted in a decreased
number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{cufflinks_asm}). The merged gene models recover 44.19\% and
58.57\% of splice junctions from ESTs and ENSEMBL respectively, which is
10-15\% greater than that from corresponding unmerged gene models.
% @CTB what about splice junctions etc??

@@

To validate our predicted isoforms, we extracted putative coding sequences from
our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan
successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
protein sequences with 50 or more amino acids.  We then searched for
homologous sequences in mouse ENSEMBL, and found that 15,399 (68.47\%)
of our isoforms from 12,399 distinct genes match mouse proteins at a
bit score $\ge1.0$.
% @CTB
These matches have a bit-score/length ratio greater than 1,
which indicates a good agreement between chicken and mouse proteins.
% @CTB why so few match, really?

@@

\caption{
\bf{Number of putative genes and isoforms}}
\begin{tabular}{cccccc}
\hline
Method& Gene & Isoform \\ 
\hline
Global & 20,439 & 26,035 \\
Local & 21,963 & 25,859 \\
Global + Local & 20,922 & 27,732 \\
Cufflinks & 25,318 & 34,959 \\
Global + Local + Cufflinks & 21,734 & 32,855 \\
Global + Local + Cufflinks + mRNAs & 26,726 & 43,074\\
Ensembl & 17,934 & 23,392 \\
\hline
%table information
\end{tabular}
%\begin{flushleft}\footnotesize \textit{*Minumum set}
%\end{flushleft}
\label{genes_transcripts}
